\chapter{Achieved results}
\label{ch:results}

% Povinné pro rigorózní práci, 3-5 stran.

\section{Symbolic computation via Program Transformation}
\label{sec:symbolic}

\subsection{Abstraction Design}

\add{ verification image }

\subsection{Term domain}

\subsection{ Program Execution }

\add{ program state }

\add{ path condition + data definitions }


\section{String Abstraction}
\label{sec:string}

\add{ aggregate abstraction }

\add{ parametric domains }

\section{Publications}
\label{sec:publications}

\begin{itemize}

\item \textbf{\emph{Symbolic Computation via Program Transformation}}, by Lauko
H., P. Ročkai and J. Barnat.  In: Theoretical Aspects of Computing -- ICTAC 2018 --
15th International Colloquium, Stellenbosch, South Africa, October 16-19, 2018,
Proceedings, pp. 313-332~\cite{Lauko2018SymComp}

\add{ I am the main author of the paper and the implementation. }

\item \emph{\textbf{Extending \textsf{DIVINE} with Symbolic Verification Using SMT}},
by \textsc{Lauko} H., V. \textsc{Štill}, P. \textsc{Ročkai} and J.\textsc{Barnat}.
In: Tools and Algorithms for the Construction and Analysis of Systems -- TACAS 2019 --
25 Years of TACAS: TOOLympics, Prague, Czech Republic, April 6-11, 2019,
Proceedings, Part III. pp. 204-208~\cite{Lauko2019Sym}

\add{ I am the main author of the paper and SMT extension. }

\item \textbf{\emph{String abstraction for model checking of C programs}},
by \textsc{Cortesi} A., H. \textsc{Lauko}, M. \textsc{Olliaro} and P. \textsc{Ročkai}.
In: Model Checking Software -- SPIN 2019 -- 26th International Symposium

\add{citation}

\add{ add description }

\end{itemize}

\section{Other Publications}

\begin{itemize}

\item \textbf{\emph{Model Checking of C and C++ with DIVINE 4}},
by Baranová Z., J. Barnat, K. Kejstová, T. Kučera, H. Lauko, J. Mrázek, P. Ročkai and V. Štill.
In: Automated Technology for Verification and Analysis -- ATVA 2017 --
15th International Symposium, Pune, India, October 3-6, 2017,
Proceedings, pp. 201-207~\cite{Divine17}

\add{ add description }

\item \emph{\textbf{Optimizing and Caching SMT Queries in SymDIVINE}},
by Mrázek J., M. Jonáš, V. Štill, H. Lauko and J. Barnat.
In: Tools and Algorithms for the Construction and Analysis of Systems -- TACAS 2017 --
23rd International Conference, Uppsala, Sweden, April 22-29, 2017,
Proceedings, Part II. pp. 390-393~\cite{Mrazek2017}

\add{ add description }

\item \textbf{\emph{\textsf{SymDIVINE}: Tool for Control-Explicit Data-Symbolic State Space Exploration}}
by Mrázek J., P. Bauch, H. Lauko and J. Barnat.
In: Model Checking Software -- SPIN 2016 -- 23rd International Symposium, Eindhoven, The Netherlands, April 7-8, 2016, Proceedings, pp. 208-213~\cite{Mrazek2016}

\add{ add description }

\end{itemize}
